Cellular Automata and Computational Universality

### 2018-08-19 paper/paper.tex
More elementary CA. Updated paper

\section{Introduction}
    A cellular automaton is a discrete model of computation which consists of a
    finite collection of ``cells'', each in one of a finite number of states.
    The state of each of these cells may evolve over the discrete progression of
    time in discrete steps, and may do so according to a deterministic set of
    rules specifying the state to which each cell is to progress, taking into
    account the states of cells in its neighbourhood. Their inherently discrete
    nature allows for strong analogies to be made with digital computers, and
    gifts them the ability to simulate digital processes and the potential to
    solve problems in this area.

@@

    Reading the bottom row of this transition table, we get the binary number
    $01011010_2 = 90_{10}$, allowing us to name this rule set Rule 90. Using
    this convention, we may now construct our own automata based solely on the
    knowledge of its name (at least for elementary cellular automata). For
    example, take Rule 110, introduced by Wolfram 1983 \cite{WolframRule110}. We
    first convert $110_{10}$ to binary, and then use this to fill in the table
    as above:

@@

    \subsection{Elementary cellular automata}
    \subsection{Life-like cellular automata}
    \subsection{Cyclic cellular automata}

>>>>

\section{Introduction}
    A cellular automaton is a discrete model of computation which consists of a
    finite collection of ``cells'', each in one of a finite number of states.
    The state of each of these cells may evolve over the progression of time in
    discrete steps, and may do so according to a deterministic set of rules
    based on the states of neighbouring cells. Their inherently discrete nature
    allows for strong analogies to be made with digital computers, and gifts
    them the ability to simulate digital processes and the potential to solve
    problems in this area.

@@

    Reading the bottom row of this transition table, we get the binary number
    $01011010_{\text{bin}} = 90_{\text{dec}}$, allowing us to name this rule set
    Rule 90. Using this convention, we may now construct our own automata based
    solely on the knowledge of its name (at least for elementary cellular
    automata). For example, take Rule 110, introduced by Wolfram 1983
    \cite{WolframRule110}. We first convert $110_{10}$ to binary, and then use
    this to fill in the table
    as above:

@@

    \subsection{Elementary cellular automata}
    Elementary cellular automata are the simplest class of one-dimensional
    cellular automata; each cell may be in one of two states, 0 or 1, and rule
    sets only take into account the cell itself and its immediate neighbours
    (the nearest cell left and right of the cell being computed). Since there
    are a total of $2^3=8$ combinations of values for a neighbourhood of cells,
    and each of these neighbourhoods causes a cell to move to one of two states
    in the next generation, this means there are $2^8=256$ elementary cellular
    automata. Their evolution is often illustrated by each progressive ``new
    line'' showing the next generation of the automaton, as caused by applying
    the rule set to the current generation.

    Two technically different automata may give rise to similar patterns, and
    indeed many automata are equivalent to each other thanks to simple
    transformations of theirunerlying geometry. The first of these
    transformations is reflection in the vertical axis; the result of applying
    this is known as the mirror rule \cite{elementary}. For example, take Rule
    30. Each time some neighbourhood produces a live cell in Rule 30, let's take
    this neighbourhood and reflect it in the vertical plane, giving the table
    below:

    \begin{table}[h]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|}\hline
            111 & 110 & 101 & 100 & 011 & 010 & 001 & 000   \\
            \hline
            0 & 0 & 0 & 1 & 1 & 1 & 1 & 0   \\
            \hline
            0 & 1 & 0 & 1 & 0 & 1 & 1 & 0   \\
            \hline
        \end{tabular}
    \end{table}

    The first line of the table shows the ruleset for Rule 30, while the line
    below gives that of our new ruleset, made by applying the mirror rule to
    Rule 30. Thus we get Rule 30's mirror rule Rule $01010110_{\text{bin}}$ =
    Rule $86_{\text{dec}}$. Rules which generate the same ruleset under
    mirroring are called amphichiral, and of the 256 elementary cellular
    automata, 64 are so.

    The second of these transformations is to exchange the roles of 0s and 1s in
    the ruleset definition, and doing so is applying the complement rule
    \cite{elementary}.  Again applying this to Rule 30, we get the table below
    (note that the roles of 0s and 1s are exchanged in both the top and bottom
    rows):

    \begin{table}[h]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|}\hline
            000 & 001 & 010 & 011 & 100 & 101 & 110 & 111   \\
            \hline
            1 & 1 & 1 & 0 & 0 & 0 & 0 & 1   \\
            \hline
        \end{tabular}
    \end{table}

    So we arrive at Rule $10000111_{\text{bin}} = 135_{\text{dec}}$ as the
    complement of Rule 30. There are 16 rules which are the same as their
    complement.
    Both of these rules can be applied at once to a ruleset, and the result is
    Rule 149 when applied to Rule 30. There are 16 rules which are the same as
    their mirrored complementary rules. There are 88 rules which are
    inequivalent under these transformations \cite{elementary}.

    \subsection{Wolfram classes}
    Wolfram gives the following classes for categorising the behaviours of
    different automata:

    \begin{itemize}
        \item Class One: rapidly converge to a uniform state e.g. Rule 0, 232
        \item Class Two: rapidly converge to a repetitive or stable state e.g.
            Rule 250
        \item Class Three: appear to remain in a random state e.g. Rule 150, 182
        \item Class Four: form areas of repetitive or stable states, but also
            structures which interact in complicated ways e.g. Rule 54
    \end{itemize}

    Two of these 256 rules are particularly special, namely Rule 54 and 110,
    both of which are Class 4 elementary cellular automata. Rule 110 has been
    proven to be universal, or Turing complete, meaning it is theorectically
    able to simulate any function. Meanwhile, it is so far unknown whether Rule
    54 is capable of universal computation -- interacting structures form, but
    it remains to been seen whether structures useful for computation can be
    formed. Computational universality is discussed more in-depth later on.

    \subsection{Two dimensional cellular automata}
    Two dimensional cellular automata become a bit more interesting than their
    dimensionally-deficient brethren. Rather than having a single row of cells
    to consider at each generation, now a two-dimensional plane of cells is used
    to compute the evolution of the automaton. Whereas elementary automata
    looked only at the neighbourhood of $a_{i-1}, a_i,$ and $a_{i+1}$ when
    applying the ruleset, two-dimensional automata have a wider choice. Two of
    the most common types of neighbourhood to consider are the Moore
    neighbourhood and the von Neumann neighbourhood. The former includes the 8
    cells directly surrounding the current cell being considered, while the
    latter only focuses on the four orthogonally ajacent cells. Ocassionally,
    one may use the extended von Neumann neighbourhood.

    \subsection{Life-like cellular automata}
    \subsection{Cyclic cellular automata}
